---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
Joshua Poirier

## Loading and preprocessing the data
If it hasn't been unzipped, unzip the *activity.zip* file from the present working directory:
```{r}
if (!file.exists("activity.csv")) {
    unzip("activity.zip")
}
```

Load the data into a data frame called *data*:
``` {r}
data <- read.csv("activity.csv", header=TRUE, nrows=17568, stringsAsFactors=FALSE)
```

Pre-process the data by formatting the *date* as a date object:
``` {r}
library(lubridate)
data$date <- ymd(data$date)
#data$time <- formatC(data$interval, width=4, format="d", flag="0")
#data$time <- paste0(substr(data$time,1,2), ":", substr(data$time,3,4))
#data$time <- hm(data$time)
```

Show some data!
``` {r echo=FALSE}
tail(data,3)
```

## What is mean total number of steps taken per day?
Build a new data frame from *data* that calculates the total number of steps taken on each day:
``` {r meanTotalSteps}
## aggregate data on 'date' field, summing the steps for each day
dailySteps <- aggregate(x=data$steps, by=list(data$date), FUN=sum, na.rm=TRUE)
names(dailySteps) <- c("Date", "TotalSteps")
meanTotalSteps <- mean(dailySteps$TotalSteps)
head(dailySteps,3)
```

The mean total number of steps taken per day is **`r round(meanTotalSteps)`**.  We show zero decimal places here as the raw data has zero decimal places, meaning zero significant digits!

``` {r echo=FALSE}
# cleanup the memory - we no longer need this data frame
rm(dailySteps)
```

## What is the average daily activity pattern?
Build a new data frame from *data* that calculates the average daily activity pattern, and call it *dailyActivity*:
``` {r avgDailyActivity}
## aggregate data on 'interval' field, finding the average steps taken for each interval
dailyActivity <- aggregate(x=data$steps, by=list(data$interval), FUN=mean, na.rm=TRUE)
names(dailyActivity) <- c("interval", "MeanSteps")
head(dailyActivity,3)
```

Let's create a time series plot of the mean steps:
``` {r echo=FALSE}
## plot the average steps taken for each interval using default plotting system
plot(dailyActivity$interval, dailyActivity$MeanSteps, type="l", 
     main="Average Daily Activity Pattern",
     xlab="Interval (hhmm)",
     ylab="Average Steps",
     xaxt="n")
axis(1, at=c(0,600,1200,1800,2400), labels=c("00:00", "06:00", "12:00", "18:00", "24:00"))
```

## Imputing missing values
In the previous step, we computed the average number of steps taken for each interval.  To impute missing values, let's insert the value from the *dailyActivity* data frame for the corresponding interval.  
``` {r message=FALSE}
library(dplyr)
## import average daily activity data frame into main data frame (linking on 'interval' field),
## then impute values from average daily activity
data <- left_join(data, dailyActivity, by="interval")
data$stepsClean <- apply(data,1,FUN=function(x) {if(is.na(x[1])) x[4] else x[1]})
print(data[2303:2311,])
```

Notice how the *stepsClean* field took the *steps* field if it was numeric, and took the *MeanSteps* field if the *steps* field was missing.  Let's also create a tidy analytic data frame for moving forward:
```{r results='hide'}
tidyData <- data.frame(date=data$date, interval=data$interval, steps=data$stepsClean)
```

## Are there differences in activity patterns between weekdays and weekends?
