---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
Joshua Poirier

## Loading and preprocessing the data
If it hasn't been unzipped, unzip the *activity.zip* file from the present working directory:
```{r}
if (!file.exists("activity.csv")) {
    unzip("activity.zip")
}
```

Load the data into a data frame called *data*:
``` {r}
data <- read.csv("activity.csv", header=TRUE, nrows=17568, stringsAsFactors=FALSE)
```

Pre-process the data by formatting the *date* as a date object:
``` {r}
library(lubridate)
data$date <- ymd(data$date)
#data$time <- formatC(data$interval, width=4, format="d", flag="0")
#data$time <- paste0(substr(data$time,1,2), ":", substr(data$time,3,4))
#data$time <- hm(data$time)
```

Show some data!
``` {r echo=FALSE}
tail(data,3)
```

## What is mean total number of steps taken per day?
Build a new data frame from *data* that calculates the total number of steps taken on each day:
``` {r meanTotalSteps}
dailySteps <- aggregate(x=data$steps, by=list(data$date), FUN=sum, na.rm=TRUE)
names(dailySteps) <- c("Date", "TotalSteps")
meanTotalSteps <- mean(dailySteps$TotalSteps)
head(dailySteps,3)
```

The mean total number of steps taken per day is **`r round(meanTotalSteps)`**.  We show zero decimal places here as the raw data has zero decimal places, meaning zero significant digits!

``` {r echo=FALSE}
# cleanup the memory - we no longer need this data frame
rm(dailySteps)
```

## What is the average daily activity pattern?
Build a new data frame from *data* that calculates the average daily activity pattern:
``` {r avgDailyActivity}
dailyActivity <- aggregate(x=data$steps, by=list(data$interval), FUN=mean, na.rm=TRUE)
names(dailyActivity) <- c("Interval", "MeanSteps")
head(dailyActivity,3)
```

Let's create a time series plot of the mean steps:
``` {r echo=FALSE}
plot(dailyActivity$Interval, dailyActivity$MeanSteps, type="l", 
     main="Average Daily Activity Pattern",
     xlab="Interval (hhmm)",
     ylab="Average Steps",
     xaxt="n")
axis(1, at=c(0,600,1200,1800,2400), labels=c("00:00", "06:00", "12:00", "18:00", "24:00"))
```

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
