---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
Joshua Poirier

## Loading and preprocessing the data
If it hasn't been unzipped, unzip the *activity.zip* file from the present working directory:
```{r}
if (!file.exists("activity.csv")) {
    unzip("activity.zip")
}
```

Load the data into a data frame called *data*:
``` {r}
data <- read.csv("activity.csv", header=TRUE, nrows=17568, stringsAsFactors=FALSE)
summary(data)
```


## What is mean total number of steps taken per day?
Build a new data frame from *data* that calculates the total number of steps taken on each day:
``` {r meanTotalSteps}
dailySteps <- aggregate(x=data$steps, by=list(data$date), FUN=sum, na.rm=TRUE)
names(dailySteps) <- c("Date", "TotalSteps")
meanTotalSteps <- mean(dailySteps$TotalSteps)
head(dailySteps,3)
```

The mean total number of steps taken per day is **`r round(meanTotalSteps)`**.  We show zero decimal places here as the raw data has zero decimal places, meaning zero significant digits!

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
